---
title: "COVID-19 Data Analysis"
date: "2024-04-04"
output:
  pdf_document: default
  html_document: default
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction:

This analysis focuses on COVID-19 data sourced from the Johns Hopkins GitHub repository. These datasets provide daily updates on the total number of confirmed COVID-19 cases and deaths for each country worldwide, as well as for each state in the United States. The datasets can be found at the following URL: https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series.

## Libraries

```{r getData, message=FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(readr)
```


## Importing Dataset and reading data from csv file

First, we need to import the required dataset from the public repository from the following source: https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/


```{r}

url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_global.csv","time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
US_cases <- read.csv(urls[3])
US_deaths <- read.csv(urls[4])

```

## TIDY AND TRANSFORM

Let us tidy the dataset and transform it into useful form so that we can use it for our further data analysis.


```{r}
global_cases <- global_cases %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>%
  select(-c(Lat,Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c('Province/State', 'Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>%
  select(-c(Lat,Long))

global <- global_cases %>% full_join(global_deaths) %>% 
  rename(Country_Region = 'Country/Region',
         Province_State = 'Province/State') %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

global <- global %>%
  unite("Combined_Key", c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

global

summary(global)

```


### Removing character "X" and converting date to date data type

Now we can convert the date to a date data type in the US data set, removing the unnecessarily long "X" character from the date. Also, we can rename the column as per our needs.

```{r tidy_US_data, message=FALSE}
US_cases <- US_cases %>% pivot_longer(cols = -(UID:Combined_Key), 
                                      names_to = "date", 
                                      values_to = "cases") %>% 
  select(Admin2:cases) %>% 
  select(-c(Lat, Long_))

US_deaths <- US_deaths %>% pivot_longer(cols = -(UID:Population), 
                                      names_to = "date", 
                                      values_to = "deaths") %>% 
  select(Admin2:deaths) %>% 
  select(-c(Lat, Long_))

US_cases$date <- gsub("X", "", as.character(US_cases$date))
US_deaths$date <- gsub("X", "", as.character(US_deaths$date))

US <- US_cases %>% full_join(US_deaths) %>% 
  mutate(date = mdy(date))
  
```

### Organizing  the data set

Now we can organize our dataset as per our analysis by grouping all the countries in a new dataframe.


```{r country}
global_country <- global %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths *1000000 / Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

global_country

```

```{r countrytotals}
total_all <- global_country %>% 
  group_by(Country_Region) %>%
  summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000*cases / population, deaths_per_thou = 1000*deaths / population) %>% filter(cases>0, population>0)

total_all %>% slice_min(deaths_per_thou, n = 10)
```

## VISUALIZATION

Data visualization is the graphical representation of information and data. First, I would like to visualize the COVID cases in each country with respect to the date.

### Covid Cases in each Country

The graph below shows the rapid rise in COVID cases in the country of Peru. Also, it shows the death rate with respect to the year.

```{r visualization}
country <- "Peru"
global_country %>% filter(Country_Region == country) %>% filter(cases > 0) %>% ggplot(aes(x=date, y=cases)) + geom_line(aes(color="cases")) + geom_point(aes(color ="cases")) + geom_line(aes(y= deaths, color="deaths")) + geom_point(aes(y=deaths, color="deaths")) + scale_y_log10() + theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("Covid Cases ", country), y=NULL)
```
Similarly, we can see the cases and deaths in Hungary and Slovakia.With this curve line, we can predict the increase and decrease in COVID cases and deaths.

```{r}
country <- "Hungary"
global_country %>% filter(Country_Region == country) %>% filter(cases > 0) %>% ggplot(aes(x=date, y=cases)) + geom_line(aes(color="cases")) + geom_point(aes(color ="cases")) + geom_line(aes(y= deaths, color="deaths")) + geom_point(aes(y=deaths, color="deaths")) + scale_y_log10() + theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("Covid Cases", country), y=NULL)
```


```{r}
country <- "Slovakia"
global_country %>% filter(Country_Region == country) %>% filter(cases > 0) %>% ggplot(aes(x=date, y=cases)) + geom_line(aes(color="cases")) + geom_point(aes(color ="cases")) + geom_line(aes(y= deaths, color="deaths")) + geom_point(aes(y=deaths, color="deaths")) + scale_y_log10() + theme(legend.position="bottom", axis.text.x = element_text(angle = 90)) + labs(title = str_c("Covid Cases in ", country), y=NULL)
```

## COVID Deaths per thousand People by Country

The below bar chart shows the COVID deaths per thousand people in each country.The below bar chart shows the COVID deaths per thousand people in each country. With this chart, we can visualize the deaths that occurred in each country.

```{r, message=FALSE}
ggplot(data = total_all, aes(x = reorder(Country_Region, +deaths_per_thou), 
                                                 y = deaths_per_thou,
                                                 fill = deaths_per_thou)) +
  scale_fill_viridis_c(option = "plasma") +
  geom_bar(stat = "identity") +
  labs(x = "Country",
      y = "Deaths per thousand People",
      title = "COVID Deaths per thousand People by Country",
      subtitle = "") +
  theme(legend.position="none", axis.text.x=element_text(angle=90, hjust=.98, vjust = .5, size=3))
```

### COVID cases in each US state

```{r more_US_data, message=FALSE}
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>% 
  summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% 
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>% 
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mil, Population) %>% 
  ungroup()
```


Let us visualize the COVID cases that occurred in each state in the US from the US dataset. The below graphs show the number of cases and deaths in each US state.

```{r}
state <- "Alabama"
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "top", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("Covid Cases in US state ", state), y = NULL)
```

```{r}
state <- "Georgia"
US_by_state %>% filter(Province_State == state) %>% filter(cases > 0) %>% 
  ggplot(aes(x = date, y = cases)) + 
  geom_line(aes(color = "cases")) + 
  geom_point(aes(color = "cases")) + 
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) + 
  scale_y_log10() + 
  theme(legend.position = "bottom", 
        axis.text.x = element_text(angle = 90)) + 
  labs(title = str_c("Covid Cases in US state ", state), y = NULL)


```
Here, we can see all the US state COVID cases together. This graph shows the number of cases in each year in each state.



```{r, message=FALSE, fig.width=10,fig.height=11}
ggplot(data = US_by_state, aes(x = date, y = cases, 
                                                   color = Province_State)) +
    geom_line() +
    facet_wrap(~Province_State) +
    labs(x = "", y = "Covid Cases", title = "COVID-19 Cases in the US",
         subtitle = "By Each state in US") +
    theme(legend.position="none")
```

## ANALYSING AND MODELING 

Let us now delve deeper into the data set by comparing COVID cases per thousand with respect to deaths per thousand. To do so, I used a linear model to show the correlation between them.
The below  model roughly follows the pandemic trend in the region. The yellow curve shows the observed deaths per thousand cases. The black one represents the linear model’s estimate of cases per thousand.


```{r}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = total_all)
summary(mod)
x_grid <- seq(1, 151)
new_df <- tibble(cases_per_thou = x_grid)
total_pred <- total_all %>% mutate(pred = predict(mod))
```

```{r}
total_pred %>% ggplot() + 
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "black") + 
  geom_point(aes(x = cases_per_thou, y = pred), color = "yellow")
```




## BIAS IDENTIFCATION:

Aggregating data from various sources introduces variability in quality, accuracy, and timeliness. The reliability of each source impacts the overall dataset. Also Inconsistent reporting of COVID cases and deaths across regions can skew the data. Varying definitions of COVID-related deaths (with or without comorbidities) add complexity. Confirmed cases may not capture all infections due to testing limitations or individual choices.

Personal Bias: Researchers and data collectors may unintentionally introduce bias based on their perspectives.
Every recorded death may be directly attributable to COVID-19. Some deaths could be due to other factors, and distinguishing them accurately can be challenging. Undiagnosed or asymptomatic cases likely went unreported. This introduces uncertainty about the actual prevalence of COVID-19.



## CONCLUSION

The data John Hopkins provides is aggregated from several different sources. It is very likely that the quality, accuracy, and timeliness of the data vary depending on the source. The analysis performed on the COVID-19 data provides valuable insights into the spread and impact of the pandemic across the globe. The visualizations and summaries generated help identify countries with the highest number of cases, deaths, and recoveries, as well as the trend in active cases over time. The visualizations and summaries generated help identify countries with the highest number of cases, deaths, and recoveries, as well as the trend in active cases over time. The global ratio of recoveries to cases indicates improving medical treatment and resilience. Despite alarming overall case and death numbers, the leveling out of cases per thousand in some countries offers hope. 

```{r session}

sessionInfo()

```
